### 2022-05-20 paper.Rmd
results / discussion

To evaluate the new k-plus objective function, three criteria were implemented in a simulation study: (i) "standard" k-plus anticlustering, minimizing differences with regard to the mean and variance; (ii) k-plus skew/kurtosis, minimizing differences with regard to the mean, variance, skew and kurtosis; (iii) k-plus correlation, minimizing differences with regard to the mean, variance and covariance structure.

@@

The same local maximum search was applied to optimize each of the competing five anticlustering methods [Method LCW in @weitz1998]. Building on an initial random assignment of elements to groups, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, for each element, all possible swaps are simulated and the resulting objective values are stored. After each exchange has been tested, the one exchange is realized that improves the objective the most. No swap is realized if no swap improves the criterion. This exchange procedure is repeated for each element. Once the last element has been processed, the procedure restarts at the beginning and once again iterates through all elements. The search stops as soon as an iteration through the entire data set no longer yielded an improvement, i.e., as soon as a local maximum has been found. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. The simulation employed five repetitions of the local maximum search.

@@

After the five anticlustering methods and the random allocation had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I compared the competing methods with regard to their ability to create similar groups. To this end, I investigated the discrepancy in means, standard deviations, skewness, kurtosis, and correlation between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$). Quantifying discrepancy in correlations ($\Delta \mathit{Cor}$) followed the same rule, but discrepancies in correlations between groups had to be aggregated across pairs of features instead of single features.

@@

As Figure 1 shows, with increasing $N$, all k-means and the k-plus methods methods converge on minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time. In particular, standard k-plus anticlustering is not much worse than k-means anticlustering when it comes to $\Delta \mathit{M}$, but it offers the immense benefit of also equalizing the spread of the data across groups.

While k-plus anticlustering apparently is capable of fulfilling multiple objectives at the same time, it is apparent that increasing the number of optimization criteria aggravates fulfilling each single criterion. For example, k-plus skew/kurtosis and k-plus correlation are worse at minimizing discrepancy with regard to standard deviations than standard k-plus. Still, every k-plus method does what it is supposed to do: k-plus skew/kurtosis is best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation is best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintain a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to be much better than diversity anticlustering with regard to skew and---in particular---kurtosis, even though this methods is specialized for these criteria. Figure 1 gives a more fine grained assessment of this observation by splitting the results by $N$: Surprisingly, for small $N$, k-plus skew/kurtosis is even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperforms diversity anticlustering when it comes to $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. These results show the cost of optimizing several criteria at the same time; increasing $N$ facilitates fulfilling multiple criteria. For low $N$, all k-plus methods show the highest discrepancies in skew and kurtosis; here k-plus seems to be preoccupied with fulfilling the other objectives. But even for small $N$, all k-plus methods outperform all other methods with regard to discrepancies in standard deviations.

Generally, it should be noted that diversity anticlustering is a good all-arounder method that tends to somewhat address all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering is clearly outperformed. As as side note, apart from k-plus correlation, diversity anticlustering is the only method that equalizes correlation structure among groups. As far as I know, this observation has not been made previously, even though the maximum diversity problem is quite well studied in the literature.

@@

>>>>

To evaluate the new family of k-plus objectives, three specific criteria were implemented in a simulation study: (i) "standard" k-plus anticlustering, minimizing differences with regard to the mean and variance; (ii) k-plus skew/kurtosis, minimizing differences with regard to the mean, variance, skew and kurtosis; (iii) k-plus correlation, minimizing differences with regard to the mean, variance and covariance structure.

@@

The same local maximum search was applied to optimize each of the five competing anticlustering methods [Method LCW in @weitz1998]. Building on an initial random assignment of elements to groups, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, for each element, all possible swaps are simulated and the resulting objective values are stored. After each exchange has been tested, the one exchange is realized that improves the objective the most. No swap is realized if no swap improves the criterion. This exchange procedure is repeated for each element. Once the last element has been processed, the procedure restarts at the beginning and once again iterates through all elements. The search stops as soon as an iteration through the entire data set no longer yielded an improvement, i.e., as soon as a local maximum has been found. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. The simulation employed five repetitions of the local maximum search.

@@

After the five anticlustering methods and the random allocation procedure had been applied to the 10,000 data sets for $K = 2$, $K = 3$, $K = 4$, I investigated how well the competing methods were able to create similar groups. To this end, I investigated the discrepancy in means, standard deviations, skewness, kurtosis, and correlation between groups---for each data set for each $K$. To quantify the discrepancy in group means, the following computation was used: for each of the (2, 3, 4 or 5) features, all (2, 3 or 4) group means were computed. For each feature, the difference between the minimum and maximum group mean was used as a measure of discrepancy; the global discrepancy in means ($\Delta M$) was then determined as the average discrepancy across features. The same procedure was applied to compute the discrepancy in standard deviations ($\Delta \mathit{SD}$), skewness ($\Delta \mathit{Skew}$) and kurtosis ($\Delta \mathit{Kurtosis}$). Quantifying discrepancy in correlations ($\Delta \mathit{Cor}$) followed the same rule, but discrepancies in correlations between groups had to be aggregated across pairs of features instead of single features.

@@

As Figure 1 shows, with increasing $N$, k-means and all k-plus methods methods converge on minimizing $\Delta \mathit{M}$. This is an important observation because it shows that including the k-plus term in the optimization process does not necessarily prevent minimizing discrepancy in means; instead, multiple objectives can be addressed at the same time. In particular, standard k-plus anticlustering is not much worse than k-means anticlustering when it comes to $\Delta \mathit{M}$, but it offers the immense benefit of also equalizing the spread of the data across groups.

While k-plus anticlustering is capable of fulfilling multiple objectives at the same time, it is apparent that increasing the number of optimization criteria aggravates fulfilling each single criterion. For example, k-plus skew/kurtosis and k-plus correlation are worse at minimizing discrepancy with regard to standard deviations than standard k-plus. Still, every k-plus method does what it is supposed to do: k-plus skew/kurtosis is best at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$, k-plus correlation is best at minimizing discrepancy with regard to $\Delta \mathit{Cor}$. At the same time, these two k-plus objectives maintain a comparably good level of addressing $\Delta \mathit{M}$ and $\Delta \mathit{SD}$, outperforming diversity anticlustering.

It is maybe surprising that k-plus skew/kurtosis does not seem to be much better than diversity anticlustering with regard to skew and---in particular---kurtosis, even though this method is specifically tailored to these criteria. Figure 1 gives a more fine grained assessment of this observation by splitting the results by $N$: Surprisingly, for small $N$, k-plus skew/kurtosis is even worse than diversity anticlustering at minimizing $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. However, with increasing $N$, it clearly outperforms diversity anticlustering when it comes to $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$. These results show the cost of optimizing several criteria at the same time; for low $N$, k-plus seems skew/kurtosis to be preoccupied with fulfilling the objectives $\Delta \mathit{M}$ and $\Delta \mathit{SD}$. Increasing $N$ then facilitates to also address $\Delta \mathit{Skew}$ and $\Delta \mathit{Kurtosis}$.

Generally, it should be noted that diversity anticlustering is a good all-arounder method that tends to somewhat address all distribution characteristics. For the specific objectives that are addressed by the k-plus methods, however, diversity anticlustering gets outperformed. As as side note, apart from k-plus correlation, diversity anticlustering is the only method that equalizes correlation structure among groups. As far as I know, this observation has not been made previously, even though the maximum diversity problem is quite well studied in the literature.

@@

Moreover, it was of interest to investigate the value of the k-plus objective function per se, independently from the algorithm that is used to optimize it. Hence, to obtain a fair assessment, it was important in the simulation study that each anticlustering objective is optimized using the same procedure, instead of using specialized multicriterion optimization schemes. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should usually be preferred because k-means does not equalize the spread of the distribution between groups, whereas the diversity maintains an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion regains a lot of attractivity: when optimizing for similarity in means and variance, k-plus anticlustering clearly outperformed the diversity criterion in the simulation study. It should also be noted that the k-plus criterion should almost always be preferred to standard k-means. This is because even for low $N$, k-plus equalizes means quite well, and for larger $N$ the practical differences between k-means and k-plus vanish completely. A reasonable exception is given when the data set consists of binary variables where mean and standard deviation are directly related.

[^clusterediting]: In their study, they used the term anticluster editing to refer to the maximization of the diversity criterion. The diversity criterion is the objective function used in the clustering method cluster editing, where it has to be minimized [@zahn1964; @shamir2004vf].
